Title

### 2016-04-25 paper/sections/ShallowOO.tex
Text for S2

@@

\begin{verbatim}
{-
type Exp = Int

lit x = x

add e1 e2 = e1 + e2
-}

{-
newtype Exp = E Int

lit x = E x

add (E e1) (E e2) = E (e1 + e2)
-}

data Exp = E {
  eval :: Int,
  transform :: Exp
}

lit x = E {eval = x, transform = lit x}

add e1 e2 = E {
  eval = eval e1 + eval e2,
  transform = add (lit 0) e2
}

p = eval $ transform (add (lit 5) (lit 3))
\end{verbatim}

>>>>

\begin{comment}
Weixin writes this part.

@@

Shallow embedding interpreter pattern
Deep embedding visitor pattern
\end{comment}

In this section, we argue that \emph{shallow embedding} and straightforward OO
programming are essentially the same thing. We use a subset
of the DSL presented in Gibbons and Wu's paper~\cite{} as the running example.
We first show how to use shallow embedding to define it and then gradually
rewrite the definition into an ``OO style''. And we also show how to do
transformations in shallow embedding, which is typically thought can only be
done in deep embedding.

\subsection{A DSL for Parallel Prefix Circuits}
Consider an DSL named \dsl that models parallel circuits.
Its BNF grammar is given below:
$$
\begin{array}{lcl}
Circuit & ::= & Identity \\
 & \mid & Fan \\
 & \mid & Beside \\
Fan & ::= & ``fan''\; PositiveNumber \\
Identity & ::= & ``identity''\;  PositiveNumber \\
Beside & ::= & Circuit ``beside'' Circuit \\
PositiveNumber & ::= & [1-9] \\
     & \mid & [1-9][0-9]^* \\
\end{array}
$$
% bnf
\weixin{the bnf grammar needs polishing}

There are three constructs - two primitives
\emph{fan} and \emph{identity} and one combinator \emph{beside} whose
meanings are: \emph{identity n} contains \emph{n} isolated vertical wires;
\emph{fan n} contains \emph{n} vertical wires and the first wire connects to
all the remaining ones from top to bottom; $beside\ c_1\ c_2$ joins two circuits
$c_1$ and $c_2$ horizontally.
For example, the circuit \emph{fan 3 beside identity 3} can be visualized in Figure~\ref{}.

\subsection{Shallow Embedding and Object-Oriented Programming}
Different from deep embedding, which uses a data structure to represent the
abstract syntax of a language, shallow embedding defines a language
directly through encoding its semantics. In the case of \dsl,
the shallow embedding implementation of language should conform to the types given below:

\lstinputlisting[language=haskell,linerange=2-5]{./code/shallowCircuit.hs}%APPLY:CTYPE
where \texttt{Circuit} models the semantic domain and should be later filled in with a
concrete type according to the semantics of the language.
For example, if the semantics of \dsl is to calculate the width of a
circuit, the definition would be:
\lstinputlisting[language=haskell,linerange=11-16]{./code/shallowCircuit.hs}%APPLY:CIRCUIT1

As shallow embedding is semantics-oriented, it is already an integer after
constructing a circuit. For example, we will immediately get $6$ for
\texttt{fan 3 `beside` identity 3} and hence \texttt{width1} is just an identity function.

The above implementation can be rewritten by wrapping the result into an
datatype and get back the value by pattern matching:
\lstinputlisting[language=haskell,linerange=20-26]{./code/shallowCircuit.hs}%APPLY:CIRCUIT2
A question then arises - is this implementation shallow or deep?

We continue rewriting the definition. This time, we use a record to represent
the result and give it a name:
\lstinputlisting[language=haskell,linerange=31-36]{./code/shallowCircuit.hs}%APPLY:CIRCUIT3
The implementation is apparently shallow since it is no different from the last version.
At this stage, however, the definition looks so like an OO program that we can easily
translate it into an OO language like Java:
\lstinputlisting[linerange=5-24]{../src/paper/sec2/Circuit.java}%APPLY:CIRCUIT_OO
The record type maps to an interface with its field declared as an abstract method.
Each case corresponds to an class and its parameters become fields of the class.
And the class implements the \texttt{Circuit} interface by giving \texttt{width}
a definition. This implementation, in fact, is an instance of the \interp pattern.
In other words, shallow embedding and the \interp are essentially the same thing.
It may also worth mentioning that deep embedding is closely related to the
\textsc{Visitor} pattern.

\subsection{Transformations in Shallow Embedding}
One of the main criticism of shallow embedding is that how to define transformations
becomes unclear since there is not a structure representing the Abstract Syntax Tree (AST).
%In fact, transformations can still be defined in shallow embedding.
For example, there usually are some language constructs that do not extend the
expressive power of the language but ease programming in the language.
Then, in implementation, we define a core language with less constructs and
define a transformation that rewrites the AST to use core constructs.

The \emph{identity n} construct in \dsl is a syntactic sugar, which can
be rewritten using the following formula:
$$identity\ n = (fan\ 1) \overbrace{`beside`\ (fan\ 1)}^{n-1}\ (n \geq 1)$$
which states that \emph{identity n} equals to \emph{n} of \emph{fan 1} combined
with \emph{beside}.
%Using the above formula, we can define a transformation that desugars every
%occurrence of \emph{identity} to a combination of \emph{fan} and \emph{beside}.

We can implement this desugaring operation in Java:
\lstinputlisting[language=haskell,linerange=29-61]{../src/paper/sec2/Circuit.java}%APPLY:DESUGAR_OO
The return type of \texttt{desugar} is \texttt{Circuit} reflecting that it is a
transformation. The implementation is straightforward: \emph{fan n} just returns
itself; $beside\ c_1\ c_2$ recursively desugars $c_1$ and $c_2$ then wrap back
their results into an instance of \texttt{Beside}; \emph{identity n} encodes the above
formula.

\weixin{define a show function in the and show the result after desugaring?}
Of course, defining \texttt{desugar} is also possible in Haskell:

\lstinputlisting[language=haskell,linerange=40-55]{./code/shallowCircuit.hs}%APPLY:DESUGAR
The definition is a little bit tricky since it makes use of Haskell's laziness.
